Monitoring forest health using hyperspectral imagery: Does feature selection improve the performance of machine-learning techniques?

### 2020-05-03 docs/00-manuscripts/ieee/ieee-manuscript.tex
optimize wording of complete manuscript

@@

\IEEEPARstart{T}{he} use of \ac{ML} algorithms for analyzing remote sensing data has seen a huge increase in the last decade~\cite{lary2016}.
This goes in line with the increased availability of remote sensing imagery, especially since the launch of the first Sentinel satellite in the year 2014.
At the same time, the implementation and usability of learning algorithms has been greatly simplified with many contributions from the open-source community.
Scientists can nowadays relatively easily process large amounts of (environmental) information using various learning algorithms.
This makes it possible to extend the matrix of possible options in a semi-automated way, possibly stumbling across unexpected findings of process settings that would have never been tested otherwise~\cite{ma2015}.

@@

Vegetation indices have shown the potential to provide valuable information when analyzing forest health \cite{jiang2014, adamczyk2015}.
Most vegetation indices were developed with the aim of being sensitive to changes of specific wavelength regions, serving as a proxy for underlying plant processes. 
However, often enough indices developed for different purposes than the one to be analyzed can help to explain complex relationships.
This emphasizes the need to extract as much information as possible from the available input data to generate promising features which can help to understand the modeled relationship.
A less known index type which can be derived from spectral information is the normalized ratio index (NRI).
In contrast to most \ac{VI} indices, the \ac{NRI} does not use an expert-based formula following environmental heuristics but instead makes use of a data-driven feature engineering approach by combining (arbitrary) combinations of spectral bands.
Especially when working with hyperspectral data, thousands of \ac{NRI} features can be derived this way.

@@

Despite its popularity in environmental modeling, there are no studies so far which used machine learning algorithms in combination with remote sensing data to analyze defoliation on a tree level.
This study aims to close this gap by analyzing defoliation at trees in northern Spain using airborne hyperspectral data.
Method-wise it makes use of machine-learning methods in combination with feature selection and hyperparameter tuning.
Incorporating the idea of creating data-driven \ac{NRI}s, this study also discusses the practical problems of high-dimensionality in environmental modeling \cite{trunk1979, xu2016}.

@@

	\item Do different environmental feature sets show differences in performance when modeling defoliation at trees?

@@

	\item Which features are most important for the models and how can these be interpreted in an ecological context?

@@

\noindent Airborne hyperspectral data with a spatial resolution of one meter and 126 spectral bands was available for four Monterey Pine (\textit{Pinus radiata}) plantations in northern Spain.
The trees in the study area plots suffer from infections of invasive pathogens such as \textit{Diplodia sapinea}, \textit{Fusarium circinatum}, \textit{Armillaria mellea} or \textit{Heterobasidion annosum}, leading to a spread of cankers or defoliation~\cite{mesanza2016, iturritxa2017}.
In-situ measurements of defoliation at trees (serving as a proxy for tree health) were collected to serve as the response variable \textit{defoliation} which ranges from 0 - 100 (in \%) (\autoref{fig:defol-distr}).
The fungi are assumed to infect the trees through open wounds, possibly caused by previous hail damage \cite{iturritxa2014}.
The dieback of these trees, which are mainly used as timber, causes high economic damages \cite{ganley2009}.

@@

\noindent The \textit{Pinus radiata} plots of this study, namely \textit{Laukiz 1}, \textit{Laukiz 2}, \textit{Luiando} and \textit{Oiartzun}, are located in the northern part of the Basque Country (\autoref{fig:study_area}).
\textit{Oiartzun} has the most observations (n = 559) while \textit{Laukiz 2} shows the largest area size (1.44 ha).
All plots besides \textit{Luiando} are located nearby the coast (\autoref{fig:study_area}).
In total 1808 observations are available (\textit{Laukiz 1} = 559, \textit{Laukiz 2} = 451, \textit{Luiando} = 301, \textit{Oiartzun} = 497).
The data was surveyed in September 2016.

@@

\noindent The airborne hyperspectral data was acquired during two flight campaigns on September 28th and October 5th 2016, both around noon.
The images were taken using a AISAEAGLE-II sensor.
All preprocessing steps (geometric, radiometric, atmospheric) were conducted by the \ac{ICGC}.
The first four bands are corrupted, leaving 122 bands with valid information.
Additional metadata information is available in \autoref{tab:hyperparameter_limits}.

@@

\subsection{Derivation of indices}
\noindent To use the full potential from the hyperspectral data, all possible vegetation indices supported by the R package $\left\{hsdar\right\}$ (89 in total) as well as all possible \ac{NRI} combinations were calculated.
The following formula was used for the NRI calculation:

@@

\noindent To account for geometric offsets (which were reported with up to 1 m from \ac{ICGC}) within the hyperspectral data, a buffer of two meters around the centroid of each tree was used for extraction of the reflectance values.
These averaged reflectance values were used as the base to derive all feature sets.
A pixel was considered to fall into a tree's buffer zone if the centroid of the respective pixel was touched by the buffer.
In total, \(\frac{121*122}{2} = 7471\) NRIs were calculated due to the first four bands of the sensor being corrupted.

@@

To reduce the dimensions of a dataset, conceptually different approaches exist: wrapper methods, filters, penalization methods (lasso and ridge) or approaches like \ac{PCA} \cite{das2001, jolliffe2016}.
Since the main focus of this study is on investigating the effect of filter methods on predictive performance, only this sub-group will be introduced in more detail in the following.

@@

% Filter methods
The concept of filters originates from the idea of ranking features using certain heuristics of an algorithm \cite{chandrashekar2014}.
Some filter methods are restricted towards specific types of variables (numeric or nominal).
Filters do only rank features, they do not choose which covariates to drop or keep \cite{drotar2015}.
The selection of features to keep for the model fitting step is usually done within the optimization phase of the model fitting, along with the hyperparameter tuning.
Essentially, the number of covariates in the model is treated as a hyperparameter of the model.
The goal is to optimize the number of features (using the ranked covariates) at which the model achieves the best performance.

% Ensemble filter methods
Besides the concept of choosing a specific filter method to rank variables, studies showed that combining several filters using statistical operations such as 'minimum' or 'mean' are able to enhance the predictive performance of the resulting models \cite{abeel2010, drotar2017a}.
This approach is referred to as 'ensemble filtering' \cite{dietterich2000}.
Ensemble filters align with the recent rise of the 'ensemble' approach in machine learning which uses stacking to combine the predictions of multiple models, aiming to enhance predictive performance \cite{polikar2012, feurer2015, bolon-canedo2019}.
In this work the 'Borda' ensemble filter was applied \cite{drotar2017a}.
For this filter, the final order is the sum of all single filters.

% Ensuring a fair weighting in the ensemble
Filter methods can be grouped into classes: Correlation-based, entropy-based, linear and non-linear methods.
It is important to not give certain classes too much weight in the ensemble as otherwise the final result will be biased towards these.
In this study this was taken care of by checking the rank correlations (Spearman's correlation) of the generated feature rankings of all methods against each other.
In case pairs of filters with a correlation above 0.9 were discovered, only one of these was included into the ensemble filter.
By this we ensured that the ensemble filter is not biased towards a certain group of methods yielding highly similar rankings.

@@

\begin{itemize}
	\item univariate/multivariate (scoring based on a single variable / multiple variables)
	\item linear/non-linear (calculation of linear/non-linear interaction terms)
	\item entropy/correlation (scoring based on derivations of entropy or correlation-based approaches)
\end{itemize}

@@

\noindent Filter 'Information Gain' is only defined for nominal response variables:

@@

where \(H\) is the conditional entropy of the response variable (class) or the feature (attribute), respectively.
To be able to use this method with a numeric response (defoliation in our case), the variable is discretized into equal bins and treated as a class variable.
While the number of bins can be treated as a hyperparameter of the filter method, we decided to use \texttt{\(n_{bin}\) = 10} after rank correlations of $> 0.9$ for different bin sizes were observed during data exploration.

@@

\noindent The benchmarking matrix of this study consists of the following algorithms:

@@

\noindent \ac{RF} and {SVM} are well established algorithms that are widely used in environmental modeling.
\ac{XGBoost} (commonly abbreviated as XGBoost) showed promising results in benchmark competitions in recent years.
Penalized regression is a statistical modeling technique capable of dealing with highly-correlated covariates by applying a penalization term which shrinks the coefficients of the model\cite{hastie2001}.
Common penalties are 'lasso' (L1) and 'ridge' (L2).
The latter does not allow the full removal of variables from the model (penalization to zero) while the former does.
Both penalties can also be combined.
The combined approach is called 'elastic net' but was not used in this work.

\subsubsection{Feature sets}
% TODO: talk about the reasons behind different features sets: environmental implications

\noindent Three feature sets were used in this study with each representing a different way of feature engineering:

\begin{itemize}
	\item The raw hyperspectral band information (HR): No feature engineering) %chktex 13
	\item Vegetation Indices (\ac{VIs}): Expert-based feature engineering)
	\item Normalized Ratio Indices (\ac{NRIs}): Data-driven feature engineering)
\end{itemize}

The idea of splitting the features into different sets originated from the question whether feature engineered indices from reflectance values have a positive effect on model performance, as for example demonstrated previously in a spectrotemporal setting \cite{pena2017}.
Benchmarking all models on these distinct groups of features makes it easier to draw conclusions on their impact when keeping all other variables such as model type, tuning strategy and partitioning method constant.
However, rather than only looking at these three groups we decided to also take their combinations into account for the overall comparison:

@@

Even though the feature selection step was passed onto the filter methods in this study, we ensured to not include features with a pairwise correlation of 1.
Having such can cause undesired effects during model fitting and feature importance calculation.
To account for such, all pair-wise correlations between features were calculated.
For pairs which exceeded the threshold of 0.9999999999 the one with the largest mean absolute correlation was removed.
This process was repeated $p$ times, each time calculating a new correlation matrix.

This preprocessing step reduced the feature sets to 122 HR features, 86 VIs and 7467 NRIs.

@@

\noindent Hyperparameters were tuned using \ac{MBO} within a nested spatial CV \cite{mlrmbo, schratz2019}.
This approach first composes \textit{n} randomly chosen hyperparameter settings out of a user defined search space.
After these \textit{n} tries have been evaluated, a new hyperparameter setting, going to be evaluated next, is proposed by a fitted surrogate model (by default a kriging method).
This strategy continues until a termination criterion, defined by the user, is reached \cite{hutter2011, jones1998}.

An initial design of 30 randomly composed hyperparameter settings and a termination criterion of 70 iterations was used, resulting in a total budget of 100 evaluated hyperparameter settings per fold.
The advantage of this tuning approach is the substantial reduction of the tuning budget required to find a setting which is close to the global minimum.
\ac{MBO} may outperform methods that do not use information from previous iterations, such as random search or grid search \cite{bergstra2012}.

For the filter methods, the percentage of dataset's features was added to the models as a hyperparameter.
For \ac{PCA}, the number of principal components was tuned instead.
Random Forest hyperparameter \texttt{\(m_{try}\)} was modified into power transformed \(p^{m_{try}}\) (ranging from 1 to $p$) to work on relative values of the respective dataset's feature count.
Here \(p\) is the number of features of the dataset.
This was necessary to ensure that \texttt{\(m_{try}\)} was not chosen out of bounds during tuning with respect to the dataset's feature count'
After the filtering, only a subset of features of the task is used for optimizing the model's hyperparameters.
The size of this subset is always different since the number of features is also optimized'
Hence, the value of \texttt{\(m_{try}\)} needs to be created dynamically based on the dataset's feature count.

@@

\noindent A spatial nested cross-validation on the plot level was chosen to reduce the influence of spatial autocorrelation as much as possible \cite{schratz2019, sperrorest}.
The \ac{RMSE} was chosen as the error measure.
Each plot served as one fold within the cross-validation setting, resulting in four iterations in total.
For the inner level (hyperparameter tuning), \(k - 1\) folds were used where \(k\) is the number of plots.

In total the benchmarking matrix consisted of 156 experiments (6 feature sets $\times$ 3 ML algorithms $\times$ 8 feature-selection methods + 2 $\times$ 6 L1 and L2 models).

@@

\noindent Estimating feature importance for highly-correlated datasets is a complicated task.
The correlation between covariates makes it challenging to calculate an unbiased estimate for single features.
Methods like \ac{PDP} or permutation-based approaches do not produce reliable estimates in such scenarios because unrealistic combinations between covariates are created \cite{molnar2019}.
Permutation feature importance calculates the loss of predictive performance for each feature by permuting it in a random manner.
The more important the feature is for the model, the higher the mean loss in the measure, reflecting the importance of the specific feature for the current model.
Predicting and calculating the error measure multiple times per feature comes with some computational cost.
For increased efficiency, we estimated the feature importance only for datasets HR and VI using the best performing learner of the benchmark.

However, as of today there are no completely satisfactory approaches for estimating feature importance for (highly) correlated datasets.
Nevertheless, recent methods like \ac{ALE} plots aim to tackle this problem by utilizing marginal distributions of features over a small window to focus on the effect of a single feature \cite{apley2019, molnar2019}.
However, ALE plots need visual inspection and as the number of features increases, making objective comparisons becomes more complicated.

In this work we applied permutation-based feature importance and \ac{ALE} plots to calculate feature importance for the HR dataset, with a focus on the former.
With the limitations of both methods in mind we aimed to get a general overview of the feature importance of the hyperspectral bands while trying to avoid the over-interpretation of results.
We used the best-performing algorithm from the HR task (i.e. SVM) to calculate the feature importance.
The ALE plots show a small relative change compared to their respective mean effects for all chosen features ($< \mid 0.001 \mid$) (\autoref{fig:fi-hr-ale}).
Most ALE curves show a high dynamic across the value range of each variable, within their respective effect range.

@@

\noindent For ecological interpretation purposes we linked the ten most important indices of the winning models for each feature set to the spectral regions of the hyperspectral data.
For feature set HR and NRI a direct linking to the respective bands of the hyperspectral sensor was done.
For the vegetation indices all bands covered by the spectral range of calculated vegetation indices were counted and summed up.

@@

\noindent The complete study was done using the open-source statistical programming language R \cite{rcoreteam2019}.
All R packages used in this study can be found in linked repositories.
Due to space limitations we will only mention the packages of the used algorithms and filter methods.

The algorithm implementations of the following packages have been used: \textit{xgboost}\cite{chen2016} (\textit{xgboost}), \textit{kernlab}\cite{kernlab} (Support Vector Machine) and \textit{glmnet}\cite{glmnet} (ridge regression).
The filter implementations of the following packages have been used: \textit{praznik\cite{praznik}}, \textit{FSelectorRcpp}\cite{fselectorrcpp}.
The R package \textit{mlr} \cite{mlr} was used for all modeling related steps.
\textit{drake} \cite{drake} was used for structuring the work and reproducibility.
This study is available as a research compendium on Zenodo (\url{10.5281/zenodo.2635403}).
The code base is available on GitHub (\url{https://github.com/pat-s/2019-feature-selection}).

@@

% - RMSE around 27
% - MBO tuned penalized methods are substantially worse
\noindent Overall, the response variable 'tree defoliation' could be modeled with an \ac{RMSE} of 28 \% percentage points (p.p.).
SVM showed no differences in RMSE across feature sets whereas other learners (RF, SVM, XGBoost, lasso and ridge) differed up to seven percentage points (\autoref{fig:perf-result}).
Ridge faced major issues in four tasks due to one observation that was predicted off the scale.
SVM showed the best performance of all learners with a difference of around three percentage points to the next best model (RF) (\autoref{tab:best-learner-perf}).
Performance differences between test folds were large: Predicting on Luiando resulted in an RMSE of 9.0 for learner SVM (without filter) but up to 54.26 p.p. when testing on Laukiz2 (\autoref{tab:svm-single-fold-perf}).

% tasks
% - no substantial difference between feature sets
% - combining feature sets has only small impact
% - expert based feature sets show no better result than data-driven ones
% - penalized methods seem to have problems with VI feature set
The combination of feature sets showed small increases in performance for some learners.
RF and XGBoost scored slightly better on the combined datasets HR-NRI and NRI-VI, respectively, compared to their standalone variants (HR, NRI, VI) (\autoref{fig:perf-result}).
Datasets containing derived features only (VI, NRI) showed no improvement in performance compared to the raw hyperspectral band information (HR).
All learners besides SVM show a substantially worse performance when using only the VI dataset (around five percentage points worse than their respective best performance).

% learner
% - SVM Car is the overall winner
% - Ridge-CV hit and miss (hit = NRI, miss = VI)
% - all MBO penalized methods have no difference across tasks
SVM combined with the 'Carscore' filter achieved the best overall performance (RMSE of 27.98 p.p.) (\autoref{tab:perf-top-10}).
%L1 penalized regression showed slightly better results in three out of six tasks when being optimized using the package internal 10-fold CV compared to MBO (RMSE of 50 vs 58).
Regression with ridge penalty (L2) showed a high variance when comparing results across tasks: In four out of six tasks (all which VI variables and HR-NRI) the error was out of bounds (\autoref{tab:perf-worst-10}).
For NRI and HR RMSE was 31.16 and 35.45 p.p., respectively.
In all settings for which ridge showed such a high error, only one observation in one fold was predicted which such a high value (in the millions).
The specific observation showed no obvious signs of being anomalous, e.g. extreme values in individual features or in principal components of features.
This one outlier caused the error estimates of these folds and the average estimate across all folds to be off the scale.

% filters
% - influence varies based on algorithm and task (SVM no influence, RF medium, XGB medium)
% - no substantial decrease in performance when using filters
% - no advantage of ensemble borda filter compared to others
Effects of filter methods on performance differed greatly between algorithms:
SVM showed no variation in performance across filters (\autoref{fig:filter-effects-no-filter}).
Using filters for RF showed a substantial increase in performance for all tasks with the exception of VI, for which the difference among all filters was also the smallest (\autoref{fig:filter-effects-no-filter}).
XGBoost showed a high dependency on filtering the data: In 4 out of 6 tasks using no filter would results in the worst or second worst performance.
However, in contrast for dataset NRI using no filter results in the best performance.
XGBoost shows the highest overall differences between filters for a single task:
For feature set HR, the range is up to 14 percentage points (Carscore vs. no filter)(\autoref{fig:filter-effects-no-filter}).

% filter vs. no filter
% - Substantial effect for XGBoost and RF
% - SVM unaffected
When comparing the usage of filters against using no filter at all, there was only one instance (XGBoost on the NRI task) when the model without filtering showed a slighly better performance than the best filtered one (\autoref{fig:filter-effects-no-filter}).
For SVM, all filters and 'no filter' achieved the same performance on tasks VI and NRI even though \autoref{fig:perf-result} lists "No Filter" as the best option.

>>>>

\usepackage{csquotes}

@@

\IEEEPARstart{T}{he} use of \ac{ML} algorithms for analyzing remote sensing data has seen a huge increase in the last decade~ cite{lary2016}.
This goes in line with the increased availability of remote sensing imagery, especially since the launch of the first Sentinel satellite in the year 2014.
At the same time, the implementation and usability of learning algorithms has been greatly simplified with many contributions from the open-source community.
Scientists can nowadays relatively easily process large amounts of (environmental) information using various learning algorithms.
This makes it possible to extend the benchmark comparison matrix of studies in a semi-automated way, possibly stumbling across unexpected findings of process settings that would have never been tested otherwise \cite{ma2015}.

@@

Vegetation indices have shown the potential to provide valuable information when analyzing forest health \cite{jiang2014, adamczyk2015}.
Most vegetation indices were developed with the aim of being sensitive to changes of specific wavelength regions, serving as a proxy for underlying plant processes. 
However, often enough indices developed for different purposes than the one to be analyzed can help to explain complex relationships.
This emphasizes the need to extract as much information as possible from the available input data to generate promising features which can help to understand the modeled relationship.
A less known index type which can be derived from spectral information is the \ac{NRI}.
In contrast to most vegetation indices, \ac{NRI}s do not use an expert-based formula following environmental heuristics but instead makes use of a data-driven feature engineering approach by combining (arbitrary) combinations of spectral bands.
Especially when working with hyperspectral data, thousands of \ac{NRI} features can be derived this way.

@@

Despite its popularity in environmental modeling, there are no studies so far which used machine-learning algorithms in combination with remote sensing data to analyze defoliation on a tree level.
This study aims to close this gap by analyzing defoliation at trees in northern Spain using airborne hyperspectral data.
The methodology of this study uses of machine-learning methods in combination with feature selection and hyperparameter tuning.
Incorporating the idea of creating data-driven \ac{NRI}s, this study also discusses the practical problems of high-dimensionality in environmental modeling \cite{trunk1979, xu2016}.

@@

	\item Do different (environmental) feature sets show differences in performance when modeling defoliation at trees?

@@

	\item Which features are most important and how can these be interpreted in an environmental context?

@@

Airborne hyperspectral data with a spatial resolution of one meter and 126 spectral bands was available for four Monterey Pine (\textit{Pinus radiata}) plantations in northern Spain.
The trees in the plots suffer from infections of invasive pathogens such as \textit{Diplodia sapinea}, \textit{Fusarium circinatum}, \textit{Armillaria mellea} or \textit{Heterobasidion annosum}, leading to a spread of cankers or defoliation~\cite{mesanza2016, iturritxa2017}.
In-situ measurements of defoliation at trees (serving as a proxy for tree health) were collected to serve as the response variable \textit{defoliation} which ranges from 0 - 100 (in \%) (\autoref{fig:defol-distr}).
It is assumed that the fungi infect the trees through open wounds, possibly caused by previous hail damage \cite{iturritxa2014}.
The dieback of these trees, which are mainly used as timber, causes high economic damages \cite{ganley2009}.

@@

The \textit{Pinus radiata} plots of this study, namely Laukiz1, Laukiz2, Luiando and Oiartzun, are located in the northern part of the Basque Country (\autoref{fig:study_area}).
Oiartzun has the most observations (n = 559) while Laukiz2 shows the largest area size (1.44 ha).
All plots besides Luiando are located nearby the coast (\autoref{fig:study_area}).
In total 1808 observations are available Laukiz1 = 559, Laukiz2 = 451, Luiando = 301, Oiartzun = 497).
The data was surveyed in September 2016.

@@

The airborne hyperspectral data was acquired during two flight campaigns on September 28th and October 5th 2016, both around noon.
The images were taken by an AISAEAGLE-II sensor.
All preprocessing steps (geometric, radiometric, atmospheric) were conducted by the \ac{ICGC}.
The first four bands were corrupted, leaving 122 bands with valid information.
Additional metadata information is available in \autoref{tab:hyperparameter_limits}.

@@

\subsection{Derivation of indices}
\noindent To use the hyperspectral data's full potential, all possible vegetation indices supported by the R package hsdar (89 in total) as well as all possible \ac{NRI} combinations were calculated from the reflectances.
The following formula was used for the NRI calculation:

@@

\noindent To account for geometric offsets within the hyperspectral data, which were reported with up to 1 m from \ac{ICGC}, a buffer of two meters around the centroid of each tree was used when extracting the reflectance values.
A pixel was considered to fall into a tree's buffer zone if the centroid of the respective pixel was touched by the buffer.
All of those pixels formed the final reflectance value of a single tree and were used as the base information to derive all additional feature sets.
In total, \(\frac{121*122}{2} = 7471\) NRIs were calculated due to the first four bands of the sensor being corrupted.

@@

To reduce the feature space of a dataset, conceptually differing approaches exist: wrapper methods, filters, penalization methods (lasso and ridge) or \ac{PCA} \cite{das2001, jolliffe2016}.
Since this study focuses mainly on filter methods, only this sub-group will be introduced in greater detail in the following subsections.

@@

% Filter methods
The concept of filters originates from the idea of ranking features using certain heuristics of an algorithm \cite{chandrashekar2014}.
Some filter methods can only deal with specific types of variables (numeric or nominal).
Filters only rank features, they do not decide which covariates to drop or keep \cite{drotar2015}.
The selection which features to keep for model fitting is usually done within the optimization phase of the model fitting, along with the hyperparameter tuning.
Essentially, the number of covariates in the model is treated as a hyperparameter of the model.
The goal is to optimize the number of features, after ranking was done, to the point at which the model achieves the best performance.

% Ensemble filter methods
Besides the concept of choosing a specific filter method to rank variables, studies showed that combining several filters using statistical operations such as 'minimum' or 'mean' are able to enhance the predictive performance of the resulting models, especially when applied to multiple datasets \cite{abeel2010, drotar2017a}.
This approach is referred to as 'ensemble filtering' \cite{dietterich2000}.
Ensemble filters align with the recent rise of the 'ensemble' approach in machine learning which uses the idea of stacking to combine the predictions of multiple models, aiming to enhance predictive performance \cite{polikar2012, feurer2015, bolon-canedo2019}.
In this work the 'Borda' ensemble filter was applied \cite{drotar2017a}.
For this filter, the final feature order is the sum of all single filters ranks.

% Ensuring a fair weighting in the ensemble
Filter methods can be grouped into classes: Correlation-based, entropy-based, linear and non-linear methods.
Care needs to be taken to not weight certain classes more than others in the ensemble as otherwise the final ranking result will be biased.
In this study this was taken care of by checking the rank correlations (Spearman's correlation) of the generated feature rankings of all methods against each other.
If filter pairs showed a correlation of 0.9 or higher, only one of the two was included into the ensemble filter, selected at random.
By this it was ensured that the ensemble filter composition was not biased towards a certain group of filter methods.

@@

\begin{itemize}
	\item Univariate/multivariate (scoring based on a single variable / multiple variables).
	\item Linear/non-linear (calculation of linear/non-linear interaction terms).
	\item Entropy/correlation (scoring based on derivations of entropy or correlation-based approaches).
\end{itemize}

@@

The filter 'Information Gain' is only defined for nominal response variables:

@@

where \(H\) is the conditional entropy of the response variable (class) or the feature (attribute), respectively.
In order to use this method with a numeric response (percentage defoliation at trees), the variable was discretized into equal bins and treated as a class variable.
While the number of bins could be treated as a hyperparameter of the filter method, it was decided to use \texttt{\(n_{bin}\) = 10} after rank correlations of $> 0.9$ for different bin sizes were observed during data exploration.

@@

The following learners were used in this work:

@@

Random forest and {SVM} are well established algorithms widely used in (environmental) modeling.
Extreme Gradient Boosting (commonly abbreviated as XGBoost) showed promising results in benchmark studies in recent years.
Penalized regression is a statistical modeling technique capable of dealing with highly-correlated covariates by penalizing the coefficients of the model \cite{hastie2001}.
Common penalties are 'lasso' (L1) and 'ridge' (L2).
Ridge does not remove variables from the model (penalization to zero) but just shrinks them to effectively zero, keeping them in the model.
The combination of both penalities is called 'elastic net' but was not used in this work.

\subsubsection{Feature sets}

Three feature sets were used in this study with each representing a different way of feature engineering:

\begin{itemize}
	\item The raw hyperspectral band information (HR): no feature engineering) %chktex 13
	\item Vegetation Indices (\ac{VI}s): expert-based feature engineering)
	\item Normalized Ratio Indices (\ac{NRI}s): data-driven feature engineering)
\end{itemize}

The idea of splitting the features into different sets originated from the question whether feature-engineered indices from reflectance values have a positive effect on model performance, as for example demonstrated previously in a spectro-temporal setting \cite{pena2017}.
Benchmarking learners on these feature sets while keeping all other variables such as model type, tuning strategy and partitioning method constant makes it possible to draw conclusions on their individual impact.
However, rather than only looking at these three groups also combinations of such were taken into account:

@@

Even though the feature-selection step should be solely left to the filter methods in this study,it was ensured a priori to account for features with a pairwise correlation of 1.
Having such features within the data can cause undesired effects during model fitting and feature importance calculation.
Hence, after having calculated all pair-wise correlations between features, for pairs which exceeded the threshold of 0.9999999999, the feature with the largest mean absolute correlation was removed from the dataset.
This process was repeated $p$ times, each time calculating a fresh correlation matrix.

This preprocessing step reduced the amount of covariates to 122 (HR), 86 (VI) and 7467 (NRI).

@@

Hyperparameters were tuned using \ac{MBO} within a nested spatial \ac{CV} \cite{mlrmbo, schratz2019}.
In MBO first \textit{n} randomly chosen hyperparameter settings out of a user defined search space are composed.
After these \textit{n} settings have been evaluated, one new setting, which going to be evaluated next, is proposed by a fitted surrogate model (by default a kriging method).
This strategy continues until a termination criterion, defined by the user, is reached \cite{hutter2011, jones1998}.

In this work, an initial design of 30 randomly composed hyperparameter settings in combination with a termination criterion of 70 iterations was used, resulting in a total budget of 100 evaluated hyperparameter settings per fold.
The advantage of this tuning approach is the substantial reduction of the tuning budget which is required to find a setting close to the global optimization minimum.
\ac{MBO} may outperform methods that do not use information from previous iterations, such as random search or grid search \cite{bergstra2012}.

To optimize the number of features used for model fitting, the percentage of features was added as a hyperparameter during the optimization stage.
For \ac{PCA}, the number of principal components was tuned instead.
The RF hyperparameter \texttt{\(m_{try}\)} was modified into power transformed \(p^{m_{try}}\) (ranging from 1 to $p$, with \(p\) being the number of features of the dataset) to work on relative values of the respective dataset's feature count.
This was necessary to ensure that \texttt{\(m_{try}\)} was not chosen higher than the available number of features left after optimizing the feature percentage during tuning.

@@

\noindent A spatial nested cross-validation on the plot level was chosen to reduce the influence of spatial autocorrelation as much as possible \cite{schratz2019, sperrorest}.
The \ac{RMSE} was chosen as the error measure.
Each plot served as one fold within the cross-validation setting, resulting in four iterations in total.
For the inner level (hyperparameter tuning), \(k - 1\) folds were used with \(k\) being the number of plots.

In total the benchmarking grid consisted of 156 experiments (6 feature sets $\times$ 3 ML algorithms $\times$ 8 feature-selection methods + 2 $\times$ 6 L1 and L2 models).

@@

\noindent Estimating feature importance for highly-correlated datasets is a complicated task.
The correlation between covariates makes it challenging to calculate an unbiased estimate for single features \cite{molnar2019}.
Methods like \ac{PDP} or permutation-based approaches do not produce reliable estimates in such scenarios because unrealistic combinations between covariates are created \cite{molnar2019}.
Permutation feature importance calculates the loss of predictive performance for each feature by permuting it in a random manner during prediction.
The more important the feature is for the model, the higher the mean loss in the measure, reflecting the importance of the specific feature for the current model.
Predicting and calculating the error measure multiple times per feature comes with some computational cost.
For increased efficiency, we estimated the feature importance only for datasets HR and VI using the best performing learner of the benchmark.

However, as of today there are no unbiased approaches for estimating feature importance for (highly) correlated datasets.
Nevertheless, recent methods like \ac{ALE} plots aim to tackle this problem by utilizing marginal distributions of features over a small window to focus on the effect of a single feature \cite{apley2019, molnar2019}.
However, ALE plots need visual inspection and as the number of features increases, making objective comparisons/conclusions becomes challenging.

In this work we applied permutation-based feature importance and \ac{ALE} plots (HR only) to estimate feature importance.
With the limitations of both methods in mind we aimed to get a general overview of the feature importance of the hyperspectral bands while trying to avoid the over-interpretation of results.
We used the best-performing algorithm on the HR task (i.e. SVM) to calculate the feature importance.

@@

For environmental interpretation purposes the ten most important indices of the best performing models of feature sets HR and VI were linked to the spectral regions of the hyperspectral data.
The aim was to visualize the most important features along the spectral curve of the plots to better understand which spectral regions were most important for the model

@@

All tasks of this study were conducted using the open-source statistical programming language R \cite{rcoreteam2019}.
A complete list of all R packages used in this study can be found in linked repositories.
Due to space limitations only the selected packages with high impact on this work will be explicitly cited.

The algorithm implementations of the following packages have been used: xgboost \cite{chen2016} (\textit{Extreme Gradient Boosting}), kernlab \cite{kernlab} (Support Vector Machine) and glmnet \cite{glmnet} (penalized regression).
The filter implementations of the following packages have been used: praznik \cite{praznik}, FSelectorRcpp \cite{fselectorrcpp}.
Package mlr \cite{mlr} was used for all modeling related steps.
drake \cite{drake} was used for structuring the work and reproducibility.
This study is available as a research compendium on Zenodo (\url{10.5281/zenodo.2635403}).
The code base is available on GitHub (\url{https://github.com/pat-s/2019-feature-selection}).

@@

% - RMSE around 27
% - MBO tuned penalized methods are substantially worse
Overall, the response variable \enquote{tree defoliation} could be modeled with an \ac{RMSE} of 28 \% percentage points (p.p.).
SVM showed no differences in RMSE across feature sets whereas other learners (RF, SVM, XGBoost, lasso and ridge) differed up to seven percentage points (\autoref{fig:perf-result}).
Ridge faced major issues in four tasks due to one observation which was predicted off the response scale (i.e. $> 100$).
SVM showed the best performance overall with a difference of around three percentage points to the next best model (RF) (\autoref{tab:best-learner-perf}).
Performance differences between test folds were large: Predicting on Luiando resulted in an RMSE of 9.0 for learner SVM (without filter) but up to 54.26 p.p. when testing on Laukiz2 (\autoref{tab:svm-single-fold-perf}).

% feature sets
% - no substantial difference between feature sets
% - combining feature sets has only small impact
% - expert based feature sets show no better result than data-driven ones
% - penalized methods seem to have problems with VI feature set
The combination of feature sets showed small increases in performance for some learners.
RF and XGBoost scored slightly better on the combined datasets HR-NRI and NRI-VI, respectively, compared to their standalone variants (HR, NRI, VI) (\autoref{fig:perf-result}).
Datasets containing derived features only (VI, NRI) showed no improvement in performance compared to the raw hyperspectral band information (HR).
All learners besides SVM showed a substantially worse performance on the VI dataset compared to all others (around five percentage points worse than their respective best performance).

% learner
% - SVM Car is the overall winner
% - Ridge-CV hit and miss (hit = NRI, miss = VI)
% - all MBO penalized methods have no difference across tasks
SVM combined with the \enquote{Carscore} filter achieved the best overall performance (RMSE of 27.98 p.p.) (\autoref{tab:perf-top-10}).
%L1 penalized regression showed slightly better results in three out of six tasks when being optimized using the package internal 10-fold CV compared to MBO (RMSE of 50 vs 58).
Regression with ridge penalty (L2) showed a high variance when comparing results across tasks: In four out of six tasks (all which VI variables and HR-NRI) the error was enourmous (\autoref{tab:perf-worst-10}).
For NRI and HR RMSE was 31.16 p.p. and 35.45 p.p., respectively.
In all settings for which ridge showed such a high error, only one observation in one fold was predicted which such a high defoliation value (in the millions).
The specific observation showed no obvious signs of being anomalous, e.g. extreme values in individual features or in principal components of features.
This one outlier caused the error estimates of these folds and the average estimate across all folds to be offscale.

% filters
% - influence varies based on algorithm and task (SVM no influence, RF medium, XGB medium)
% - no substantial decrease in performance when using filters
% - no advantage of ensemble borda filter compared to others
Effects of filter methods on performance differed greatly between algorithms:
SVM showed no variation in performance across filters (\autoref{fig:filter-effects-no-filter}).
Using filters for RF showed a substantial increase in performance for all tasks with the exception of VI, for which the difference among all filters was also the smallest (\autoref{fig:filter-effects-no-filter}).
XGBoost showed a high dependency on filtering the data: In 4 out of 6 tasks using no filter resulted in the worst or second worst performance.
In contrast, using no filter on dataset NRI resulted in the best performance.
XGBoost shows the highest overall differences between filters for a single task:
for feature set HR, the range is up to 14 percentage points (\enquote{Carscore} vs. \enquote{no filter})(\autoref{fig:filter-effects-no-filter}).

% filter vs. no filter
% - Substantial effect for XGBoost and RF
% - SVM unaffected
When comparing the usage of filters against using no filter at all, there was only one instance (XGBoost on the NRI task) when a model without filtering scored a better performance than the best filtered one (\autoref{fig:filter-effects-no-filter}).
For SVM, all filters and \enquote{no filter} achieved the same performance on tasks VI and NRI even though \autoref{fig:perf-result} lists "No Filter" as the best option.

% Borda vs. other filters
% - most often among the first best filters for each task
% - never the best
The Borda filter did not achieve the best performance for any learner across any task \autoref{fig:filter-effects-borda}.
For RF and XGBoost it most often ranked within the first 50\% with respect to all filters of a specific task.
For XGBoost on the VI task, the Borda filter scored the second worst performance.
